experDesign: stratifying samples into batches with minimal bias

### 2021-11-10 paper.Rmd
Update plots and text accordingly

---
title: 'experDesign: stratifying samples into batches with minimal bias'
output:
  html_document:
    df_print: paged
  word_document:
    keep_md: yes
  pdf_document: default
  md_document:
    preserve_yaml: yes
date: "25th August 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup}
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE, 
                      error = 0,
                      fig.width = 9)
```

@@

OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
Moreover, it works for both numeric and categorical variables:

@@

OSAT might return less rows than the input provided because they might have NA value.

@@

anticlust does not handle all the variables, it only accepts numeric variables.

@@

```

```{r panel1, echo=FALSE, message=FALSE, warning=FALSE}
(osat_smoke/(osat_age + osat_sex) / anticlust & plot_layout(guides = 'collect')
```

@@

```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with anticlust.", warning=FALSE, message=FALSE}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations, name = "")
experDesign_index_spatial <- spatial(index = experDesign_index, 
                                     pheno = survey[, VoI], 
                                     iterations = iterations, 
                                     rows = LETTERS[1:8], 
                                     columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + 
                      experDesign_sex )) +
  plot_annotation(title = "experDesign") +
  plot_layout(guides = 'collect')
```

The statistics of the index can be checked for multiple statistics, as shown on the help pages of `evaluate_na`, `evaluate_entropy`, `evaluate_mad`, `evaluate_sd` and `evaluate_mean`.

@@

We can also compare our results with the original distribution via `evaluate_orig`.

>>>>

---
title: 'experDesign: stratifying samples into batches with minimal bias'
output:
  md_document:
    preserve_yaml: yes
  word_document:
    keep_md: yes
  pdf_document: default
  html_document:
    df_print: paged
date: "11th November 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
	error = 0,
	fig.width = 9,
	warning = TRUE,
	collapse = TRUE
)
```

@@

OSAT provides templates for plates that hold 2, 4, 8 Illumina BeadChip chips, having 24, 48 or 96 wells.
Moreover, it works for both numeric and categorical variables but OSAT might return less rows than the input provided because they might have NA value.

@@

@@

anticlust does not handle all types of variables, it only accepts numeric variables.

@@

```
These methods results on this distributions:

```{r panel1, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure A: distribution of  samples according to *OSAT*, Figure B according to *anticlust* on the variables used (1: Smoke, 2: Age, 3: Sex)."}
osat_plot <- osat_smoke / (osat_age + osat_sex)
osat_plot /anticlust + plot_layout(guides = 'collect') +
  plot_annotation(tag_levels = list(c("A.1", "A.2", "A.3", "B.2")),
                  title = "Existing solutions",
                  subtitle = "OSAT and anticlust solutions") &
  theme(plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 8, hjust = 0, vjust = 0))
```

On Figure 1, we can see that the distributions is fairly similar between these two packages.

@@

```{r experDesign, echo=FALSE, fig.cap="Sample, distributions with anticlust.", warning=FALSE, message=FALSE}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations, name = "")
experDesign_index_spatial <- spatial(index = experDesign_index, 
                                     pheno = survey[, VoI], 
                                     iterations = iterations, 
                                     rows = LETTERS[1:8], 
                                     columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
# nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + 
                      experDesign_sex )) +
  plot_annotation(title = "experDesign") +
  plot_layout(guides = 'collect')
```

The statistics of the index can be checked for multiple statistics, as shown on the help pages of `evaluate_na`, `evaluate_entropy`, `evaluate_mad`, `evaluate_sd` and `evaluate_mean`. We can also compare our results with the original distribution via `evaluate_orig`.

@@
